Forcing for Isabelle

### 2018-02-14 blueprint.tex
truth lemma for atomic formulas, statement

\subsection{Forcing for atomic formulas}
The base case of the definition of the forcing relation, that is, for
atomic formulas, is (in)famously the most complicated one. 
The definition adopted by Kunen~\cite[p.~257]{kunen2011set} is the
following, with two mutually recursive clauses:
\begin{multline}\label{eq:def-forcing-equality}
  \forceisa({t_1= t_2},\PP,\leq,\1,p) \defi 
  \forall s\in\dom(t_1)\cup\dom(t_2)\;\forall q\leq p :\\
  \forceisa({s\in t_1},\PP,\leq,\1,q)\lsii 
  \forceisa({s\in t_1},\PP,\leq,\1,q),
\end{multline}
\begin{multline}\label{eq:def-forcing-membership}
  \forceisa({t_1\in t_2},\PP,\leq,\1,p) \defi  \\
  \text{``}\{q\leq p : \exists \lb s,r\rb\in t_2 (q\leq r \y
  \forceisa({t_1 =s},\PP,\leq,\1,q))\} \\
  \text{is dense below $p$.''}
\end{multline}
Other approaches, like the one in Neeman~\cite{neeman-course}, prefer
to have a single, more complicated, clause for $t_1=t_2$ using
well-founded recursion on $t_1$
and $t_2$ and then define the case $t_1\in t_2$ explicitly.

@@

\section{The fundamental theorems}
These are stated as two lemmas: \emph{Truth} and
\emph{Definability}. The second one ``simply'' says that the forcing
relation is definable. The previous recursive construction amounts to
that. The Truth Lemma states that the forcing relation indeed relates
satisfaction in $M[G]$ to that in $M$. 
\begin{lemma}[Truth Lemma]
  Let $M$ be a ctm of $\ZFC$ and $\lb\PP,\leq,\1\rb$ a forcing notion
  in $M$, $p\in\PP$, and $\phi(x_1,\dots,x_n)$ a formula in the
  language of set 
  theory with all free variables displayed. Then the
  following are equivalent, for all $\tau_1,\dots,\tau_n\in M$: 
  \begin{enumerate}
  \item     $M[G]\models\phi(\val(\tau_1,G),\dots,\val(\tau_n,G))$ for
    all filters $G$, $M$-generic for $\PP$, such that $p\in G$.
  \item $M\models\forceisa(\phi(\tau_1,\dots,\tau_n),\PP,\leq,\1,p)$.
  \end{enumerate}
\end{lemma}
A bit more formally, one should write the parameters to the left of
$\models$:
\begin{gather*}
  M[G],[\val(\tau_1,G),\dots,\val(\tau_n,G)]
  \models\phi(x_1,\dots,x_n) \\
  \text{if and only if}\\
  M,[\tau_1,\dots,\tau_n,\PP,\leq,\1,p] 
  \models\forceisa(\phi(x_1,\dots,x_n),x_{n+1},x_{n+2},x_{n+3},x_{n+4})
\end{gather*}

>>>>

\subsection{Forcing for atomic formulas}\label{sec:forcing-atomic-formulas}
The base case of the definition of the forcing relation, that is, for
atomic formulas, is (in)famously the most complicated one. 
The definition adopted by Kunen~\cite[p.~257]{kunen2011set} is the
following, with two mutually recursive clauses:
\begin{multline}\label{eq:def-forcing-equality}
  \forceisa({t_1= t_2},\PP,\leq,\1,p) \defi 
  \forall s\in\dom(t_1)\cup\dom(t_2)\;\forall q\leq p :\\
  \forceisa({s\in t_1},\PP,\leq,\1,q)\lsii 
  \forceisa({s\in t_1},\PP,\leq,\1,q),
\end{multline}
\begin{multline}\label{eq:def-forcing-membership}
  \forceisa({t_1\in t_2},\PP,\leq,\1,p) \defi  \\
  \text{``}\{q\leq p : \exists \lb s,r\rb\in t_2 (q\leq r \y
  \forceisa({t_1 =s},\PP,\leq,\1,q))\} \\
  \text{is dense below $p$.''}
\end{multline}
Other approaches, like the one in Neeman~\cite{neeman-course}, prefer
to have a single, more complicated, clause for $t_1=t_2$ using
well-founded recursion on $t_1$
and $t_2$ and then define the case $t_1\in t_2$ explicitly.

@@

\section{The fundamental theorems}
These are stated as two lemmas: \emph{Truth} and
\emph{Definability}. The second one ``simply'' says that the forcing
relation is definable. The previous recursive construction amounts to
that. The Truth Lemma states that the forcing relation indeed relates
satisfaction in $M[G]$ to that in $M$. 
\begin{lemma}[Truth Lemma]\label{lem:truth-lemma}
  Let $M$ be a ctm of $\ZFC$, $\lb\PP,\leq,\1\rb$ a forcing notion
  in $M$, $p\in\PP$, and $\phi(x_1,\dots,x_n)$ a formula in the
  language of set 
  theory with all free variables displayed. Then the
  following are equivalent, for all $\tau_1,\dots,\tau_n\in M$: 
  \begin{enumerate}
  \item     $M[G]\models\phi(\val(\tau_1,G),\dots,\val(\tau_n,G))$ for
    all filters $G$, $M$-generic for $\PP$, such that $p\in G$.
  \item $M\models\forceisa(\phi(\tau_1,\dots,\tau_n),\PP,\leq,\1,p)$.
  \end{enumerate}
\end{lemma}
A bit more formally, one should write the parameters to the left of
$\models$:
\begin{gather*}
  \forall G : p\in G \limp M[G],[\val(\tau_1,G),\dots,\val(\tau_n,G)]
  \models\phi(x_1,\dots,x_n) \\
  \text{if and only if}\\
  M,[\tau_1,\dots,\tau_n,\PP,\leq,\1,p] 
  \models\forceisa(\phi(x_1,\dots,x_n),x_{n+1},x_{n+2},x_{n+3},x_{n+4})
\end{gather*}
The proof of the Truth Lemma is by recursion in elements of
$\formula$. The following auxiliary result (adapted from
\cite[IV.2.43]{kunen2011set}) is also proved by recursion in
$\formula$.
\begin{lemma}\label{lem:3-density}
  Under the same hypothesis of Lemma~\ref{lem:truth-lemma}, we have
  \begin{enumerate}
  \item $M\models\forceisa(\phi,\PP,\leq,\1,p)$ and $p_1\leq p$
    implies $M\models\forceisa(\phi,\PP,\leq,\1,p_1)$.
  \item $M\models\forceisa(\phi,\dots)$ if and only if
    $M\models$``$\{p_1\leq p : \forceisa(\phi,\PP,\leq,\1,p_1)\}$ is
    dense below $p$.''
  \item $M\models\forceisa(\phi,\dots)$ if and only if 
    $M\models\neg\exists
  q\leq p \,(\neg \forceisa(\phi,\PP,\leq,\1,q))$.
  \end{enumerate}
\end{lemma}

\subsection{Truth for atomic formulas}
We restate Lemma~\ref{lem:truth-lemma} in an alternative way, for
atomic formulas.
\begin{lemma}
  Let $M$ be a ctm of $\ZFC$, $\lb\PP,\leq,\1\rb$ a forcing notion
  in $M$, $p\in\PP$, $G\sbq\PP$ $M$-generic for $\PP$ and $\phi(x_1,x_2)$ an
  atomic formula. Then the following hold for all
  $\tau_1,\tau_2\in M$:
  \begin{enumerate}
  \item If $p\in G$ and
    \begin{equation} \label{eq:M-sats-forces-atom}
      M,[\tau_1,\tau_2,\PP,\leq,\1,p] 
      \models\forceisa(\phi(x_1,x_2),x_{3},x_{4},x_{5},x_{6})
    \end{equation}
    then
    \begin{equation} \label{eq:MG-sats-atom}
      M[G],[\val(\tau_1,G),\val(\tau_2,G)]
      \models\phi(x_1,x_2).
    \end{equation}
  \item If (\ref{eq:MG-sats-atom}) holds 
    then there is some $p\in G$ such that
    (\ref{eq:M-sats-forces-atom}) holds for that $p$.
  \end{enumerate}
\end{lemma}
This proof consists of two cases for $\phi$ (namely, $x_1\in x_2$ and
$x_1=x_2$) and it is done by well-founded recursion over $R$ (defined
in Section~\ref{sec:forcing-atomic-formulas}).
